---
title: "Ellison Scale Tests"
output: html_notebook
---



```{r, echo = F}
library(dplyr)
library(dbplyr)
library(ggplot2)
library(gsw)
library(tidyverse)
```

```{r, echo=F}
#connect data base
vmp_profiles <- DBI::dbConnect(RSQLite::SQLite(), "../data/SQLite_VMP.sqlite")
# show infos about database
src_dbi(vmp_profiles)

# read out data from fastData table for profile 53 (KZN_032)
vmp53 <- tbl(vmp_profiles, "VMP_profiles_fastData")
vmp53_df <- vmp53 %>%
  filter(profile_nr == 53) %>%
  select(-c(Chla_mug_L, Turbidity)) %>%
  collect()

head(vmp53_df)

# sort profiles for temperature
b <- vmp53_df
# sort pressure increasing
b <- b[order(b$Pressure_dbar, decreasing=F),]
#2nd add sorted T1 column a
b_t <- select(b,Pressure_dbar,T1_degC,T2_degC)   #pressure, T1, T2
# sort T1 decreasing!
b_t1s <- b_t[order(b_t$T1_degC, decreasing=T),]
b_t2s <- b_t[order(b_t$T2_degC, decreasing=T),]
# add sorted T1 to b  
b$T1_s <- b_t1s$T1_degC
b$T2_s <- b_t2s$T2_degC
b$pressure1_s <- b_t1s$Pressure_dbar
b$pressure2_s <- b_t2s$Pressure_dbar
b$tsc1 <- b$pressure1_s - b$Pressure_dbar
b$tsc2 <- b$pressure2_s - b$Pressure_dbar
head(b)
```

## Examples of temperature profiles

Some examples of sorted (over entire depth) versus original temperature profiles for different segments, shown over _fft_length_ (1024*0.5, ~0.5m, _s in legend means sorted profile):

```{r, echo=F}
seg_length <- 1024 * 0.5

stvec <- c(20, 50, 100, 600, 5, 200, 500, 700)
c <- data.frame()
for(i in 1:length(stvec)){
  seg_st <- stvec[i] * seg_length
  seg_en <- seg_st + seg_length -1
  b_i <- b[seg_st:seg_en, ]

  ci <- b_i %>%
    select(profile_nr,Pressure_dbar, time_sec,T1_degC, T2_degC, T1_s, T2_s) %>%
    gather(key=Temp_var, -c(Pressure_dbar,time_sec,profile_nr), value=T_degC)
  ci$seg_midpr <- round(mean(ci$Pressure_dbar),0) 
  
  c <- rbind(c,ci)

}
p1 <- ggplot(data = c) + 
    geom_line(aes(y=T_degC, x=Pressure_dbar, col=Temp_var)) +
    scale_x_reverse() +
    coord_flip() +
  theme_bw() +
  facet_wrap(~ seg_midpr, nrow=2, scales='free')
p1

```

Some examples of sorted versus original temperature profiles for different segments, shown over larger range, 1024*2, ~2m:

```{r, echo=F}
seg_length <- 1024 * 2

stvec <- c(0, 1, 5, 10, 20, 50, 80, 120)
c <- data.frame()
for(i in 1:length(stvec)){
  seg_st <- stvec[i] * seg_length
  seg_en <- seg_st + seg_length -1
  b_i <- b[seg_st:seg_en, ]

  ci <- b_i %>%
    select(profile_nr,Pressure_dbar, time_sec,T1_degC, T2_degC, T1_s, T2_s) %>%
    gather(key=Temp_var, -c(Pressure_dbar,time_sec,profile_nr), value=T_degC)
  ci$seg_midpr <- round(mean(ci$Pressure_dbar),0) 
  
  c <- rbind(c,ci)

}
p2 <- ggplot(data = c) + 
    geom_line(aes(y=T_degC, x=Pressure_dbar, col=Temp_var)) +
    scale_x_reverse() +
    coord_flip() +
  theme_bw() +
  facet_wrap(~ seg_midpr, nrow=2, scales='free')
p2

```

### same profile plots for slow data (density and temperature)

```{r, echo=F}
#connect data base
vmp_profiles <- DBI::dbConnect(RSQLite::SQLite(), "../data/SQLite_VMP.sqlite")
# read out data from fastData table for profile 53 (KZN_032)
vmp53sl <- tbl(vmp_profiles, "VMP_profiles_slowData_cleaned")
vmp53sl_df <- vmp53sl %>%
  filter(profile_nr == 53) %>%
  collect()
DBI::dbDisconnect(vmp_profiles)

#head(vmp53sl_df)

# practical salinity using conductivity in ...?
vmp53sl_df$SP <- gsw_SP_from_C(vmp53sl_df$Conductivity_mS_per_cm,vmp53sl_df$Temperature_degC,vmp53sl_df$Pressure_dbar)
# absolute salinity
#average latitude for vmp profiles = -27.52277 and av long = 32.7259
long <- 32.7259
lat <- -27.52277
vmp53sl_df$SA <- gsw_SA_from_SP(vmp53sl_df$SP,vmp53sl_df$Pressure_dbar,long,lat)
# conservative temperature from in-situ temperature
vmp53sl_df$CT <- gsw_CT_from_t(vmp53sl_df$SA,vmp53sl_df$Temperature_degC,vmp53sl_df$Pressure_dbar)
# in-situ density
vmp53sl_df$rho <- gsw_rho(vmp53sl_df$SA,vmp53sl_df$CT,vmp53sl_df$Pressure_dbar)
# potential density
p_ref = 0      #[dbar] --> =0bar (next would be 100bar=1000dbar, but max p of 300dbar closer to 0
vmp53sl_df$rho_pot <- gsw_rho(vmp53sl_df$SA,vmp53sl_df$CT,p_ref)

# sort profiles for temperature
bsl <- vmp53sl_df
# sort pressure increasing
bsl <- bsl[order(bsl$Pressure_dbar, decreasing=F),]
#2nd add sorted rho/temp column a
bsl_R <- select(bsl,Pressure_dbar,rho_pot)   # potential density
bsl_T <- select(bsl,Pressure_dbar,Temperature_degC)       # temp
# sort density/temp increasing/decreasing!
bsl_Rs <- bsl_R[order(bsl_R$rho_pot, decreasing=F),]
bsl_Ts <- bsl_T[order(bsl_T$Temperature_degC, decreasing=T),]
# add sorted T1 to b  
bsl$R_s <- bsl_Rs$rho_pot
bsl$T_s <- bsl_Ts$Temperature_degC
bsl$pressureR_s <- bsl_Rs$Pressure_dbar
bsl$pressureT_s <- bsl_Ts$Pressure_dbar
bsl$tscR <- bsl$pressureR_s - bsl$Pressure_dbar
bsl$tscT <- bsl$pressureT_s - bsl$Pressure_dbar
#head(bsl)
```

```{r, echo=F}
seg_length <- 128 * 20

stvec <- c(0,1,2,4, 6,10,12,15) # 10, 20, 30, 50, 60)
# stvec <- c(0, 1, 5, 10, 20, 30, 50, 60) # for 128 * 5
csl <- data.frame()
for(i in 1:length(stvec)){
  seg_st <- stvec[i] * seg_length
  seg_en <- seg_st + seg_length -1
  b_i <- bsl[seg_st:seg_en, ]

  ci <- b_i %>%
    select(profile_nr,Pressure_dbar, time_sec, Temperature_degC, rho_pot, T_s, R_s) %>%
    gather(key=rho_var, -c(profile_nr, Pressure_dbar, time_sec, Temperature_degC, T_s), value=rho_pot) %>%
    gather(key=temp_var, -c(profile_nr, Pressure_dbar, time_sec, rho_var, rho_pot), value=temp)
    
  ci$seg_midpr <- round(mean(ci$Pressure_dbar),0) 
  
  csl <- rbind(csl,ci)

}
#plot temp
ptemp <- ggplot(data = csl) + 
  geom_line(aes(y=temp, x=Pressure_dbar, col=temp_var)) +
  scale_x_reverse() +
  coord_flip() +
  theme_bw() +
  facet_wrap(~ seg_midpr, nrow=2, scales='free')
ptemp

# plot density
prho <- ggplot(data = csl) + 
  geom_line(aes(y=rho_pot, x=Pressure_dbar, col=rho_var)) +
  scale_x_reverse() +
  coord_flip() +
  theme_bw() +
  facet_wrap(~ seg_midpr, nrow=2, scales='free')
prho

```

### Plot sorted pressure from rho and temp over actual pressure

```{r, echo=F}
seg_length <- 128 * 20

stvec <- c(0,1,2,4, 6,10,12,15) # 10, 20, 30, 50, 60)
# stvec <- c(0, 1, 5, 10, 20, 30, 50, 60) # for 128 * 5
dsl <- data.frame()
for(i in 1:length(stvec)){
  seg_st <- stvec[i] * seg_length
  seg_en <- seg_st + seg_length -1
  b_i <- bsl[seg_st:seg_en, ]

  di <- b_i %>%
    select(profile_nr,Pressure_dbar, pressureT_s, pressureR_s) %>%
    gather(key=p_var, -c(profile_nr, Pressure_dbar), value=pressure)
  
  di$seg_midpr <- round(mean(di$Pressure_dbar),0) 
  
  dsl <- rbind(dsl,di)

}
#plot sorted pressure
pp <- ggplot(data = dsl) + 
  geom_line(aes(y=pressure, x=Pressure_dbar, col=p_var)) +
  scale_x_reverse() +
  coord_flip() +
  theme_bw() +
  facet_wrap(~ seg_midpr, nrow=2, scales='free')
pp
```


### Plot of r^2 of fitted linear model for sections of 1024*0.5 data points (about 0.5 m):

```{r, echo=F}
# fit model through segments
# model data
md <- b

fs <- 1024
fft_length <- 0.5
seg_length <- fs * fft_length
overlap <- fs * fft_length/2  #seg_length/(fft_length/2)

#nr_segments <- floor(length(b$time_sec)/(seg_length - overlap))

rsq <- data.frame(p_mid = NA, rsqT1 = NA, rsqT2 = NA, rsqT1b = NA, rsqT2b = NA)
seg_st <- 1
seg_en <- seg_st + seg_length - 1
md_i <- md[seg_st:seg_en, ]
count <- 1
while(length(md$profile_nr) > (seg_st + seg_length - 1)){
  #for(i in 1:nr_segments){
  mod1 <- summary(lm(T1_s ~ Pressure_dbar, md_i))
  mod2 <- summary(lm(T2_s ~ Pressure_dbar, md_i))
  # do for sorted T per segment
  md_i_t <- select(md_i,Pressure_dbar,T1_degC,T2_degC)   #pressure, T1, T2
  # sort T1 decreasing!
  md_i_t1s <- md_i_t[order(md_i_t$T1_degC, decreasing=T),]
  md_i_t2s <- md_i_t[order(md_i_t$T2_degC, decreasing=T),]
  md_i$T1b_s <- md_i_t1s$T1_degC
  md_i$T2b_s <- md_i_t2s$T2_degC
  mod1b <- summary(lm(T1b_s ~ Pressure_dbar, md_i))
  mod2b <- summary(lm(T2b_s ~ Pressure_dbar, md_i))
  rsq_i <- data.frame(p_mid = mean(md_i$Pressure_dbar), 
                      rsqT1 = mod1$r.squared, 
                      rsqT2 = mod2$r.squared,
                      rsqT1b = mod1b$r.squared,
                      rsqT2b = mod2b$r.squared)
  rsq <- rbind(rsq, rsq_i)
  # set new segment start/end
  seg_st <- seg_st + (seg_length-overlap)
  seg_en <- seg_st + seg_length - 1
  md_i <- md[seg_st:seg_en, ]
  count <- count + 1
}
rsq <- rsq[-1,]
```


```{r, echo = F}
pl.data <- gather(rsq, key = var, -p_mid, value = rsq)
pl.data$rcat <- ifelse(pl.data$var %in% c('rsqT1','rsqT2'), 'entire profile sorted', 'segment sorted')
ggplot(data = pl.data) + 
  geom_line(aes(y = rsq, x = p_mid, col = var)) +
  scale_x_reverse() +
  coord_flip() +
  theme_bw() +
  facet_grid(. ~ rcat)
```

## Calculate Ellison Scale

Ellison scale is calculated for each data point by taking the deviation of the measured temperature from the sorted temperature and devide by the displacement (thorpe scale) resulting from the sorting process. 

```{r}
profnum <- 53
fs <- 1024
diss_length <- 5
overlap <- diss_length/2
seg_length <- fs * diss_length
overlap_length <- fs * overlap

d <- data.frame()
seg_st <- 1
seg_en <- seg_st + seg_length - 1
b_i <- b[seg_st:seg_en, ]
while(length(b$Pressure_dbar) >= (seg_st + seg_length - 1)){
  attach(b_i)
  Tdev1 <- abs(T1_degC - T1_s)
  Tdev2 <- abs(T2_degC - T2_s)
  LE1i <- ifelse(Tdev1 == 0, 0, Tdev1/abs(tsc1))
  LE2i <- ifelse(Tdev2 == 0, 0, Tdev2/abs(tsc2))
  LEi <- data.frame(profile_nr=profnum, pressure=mean(Pressure_dbar), 
                      LE_T1=sqrt(mean(LE1i^2)), LE_T2=sqrt(mean(LE2i^2)),
                      tsc_T1=sqrt(mean(b_i$tsc1^2)), tsc_T2=sqrt(mean(b_i$tsc2^2)))
  detach(b_i)
  d <- rbind(d, LEi)
  seg_st <- seg_st + (seg_length-overlap_length)
  seg_en <- seg_st + seg_length - 1
  b_i <- b[seg_st:seg_en, ]
  }
```

```{r}
ggplot(data=d, aes(x=LE_T1)) + geom_density()
ggplot(data=d, aes(x=LE_T1, y = tsc_T1)) + geom_point()
```