---
title: "Estimating chi"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

Spectra of temperature gradients are located within the diss structure created in matlab when running quick_look (find them in diss.sclar_spectra.scalar_spec). They are the uncorrected frequency spectra. Function get_scalar_spectra-odas.m calculates spectra of scalar vectors (including fast temperature gradients), uses functions clean_shear_spec.m or csd_odas.m to calculate spectra.

This notebook imports the data saved from matlab, corrects for the frequency response of the sensor using Vachon and Lueck correction (used by Rockland in odas and by Bluteau) and tries to fit the model stated in Bluteau to the spectrum for k<70 cpm.

Data from profile 032 used for testing, dive depth was 414 m. 146 spectra were calculated.

##### Following questions:
- which range of wavenumbers to fit model to? --> 70 cpm?
- data spectrum doesnt really increase? like in spectrums in Bluteau
- model doesn't seem to fit data very well, why?, do I miss some more transformation of spectrum?
- do I have to convert from cpm to rad/m?


```{r message=FALSE, warning=FALSE, include=FALSE}
library(R.matlab)
library(tidyverse)
library(ggplot2)
library(stats4)  # for mle 
```

### Reading in the data

```{r}
# read in data
diss <- readMat('../data/Diss_032_fft2.mat')
diss_fieldnames <- unlist(readMat('../data/diss_fieldnames_fft2.mat'))
# spectrum energy values
# spec 3D array
spec1 <- diss[[1]][[47]][[1]]
gradT1 <- spec1[,1,]
gradT2 <- spec1[,2,]

rm(spec1)
```

```{r}
# wavenumber 2D matrix with each column for one depth segment
waven1 <- diss[[1]][[47]][[3]]

# frequency matrix
freq <- diss[[1]][[47]][[2]]
```

```{r}
# additional variables
mean_speed <- diss[[1]][[16]]
pressure <- diss[[1]][[18]]
eps <- diss[[1]][[1]]
rm(diss)
```

### Correcting for frequency response

The following calculations are done exemplary for only one temperature sensor.

 - expression for tau from odas library

```{r}
F_0 <- 25*sqrt(mean_speed)
tau_therm <- 2*pi*F_0 / sqrt(sqrt(2) - 1)   # time constant for double-pole response
tau_therm <- 1 / tau_therm
#correction:
gradT1_c <- matrix(data=NA, nrow=length(freq[,1]), ncol=length(freq[1,]))
for(index in 1:length(freq[1,])){
  gradT1_c[,index] <- gradT1[,index] * (1 + (2*pi*tau_therm[index]*freq[,index])^2)^2
}
```

#### Plot of uncorrected and corrected spectrum

```{r echo=FALSE}
specnr <- 30
plot(log10(freq[,specnr]), log10(gradT1_c[,specnr]), type='l', ylim=c(-6,-2))
lines(log10(freq[,specnr]), log10(gradT1[,specnr]), col='red')
title('frequency spectrum corrected (black), uncorrected (red)')
```



### Fit model to spectra with chi as parameter estimated with MLE 

model to fit:
psi_gradT(k) = C_T * chi * eps^-1/3 * k^1/3

  with psi_gadT --> corrected energy spectrum,
  
  C_T = Obukhov-Corrsin universal constant with values between 0.3 and 0.5, 0.4 recommended,
  
  and k = wavenumber

#### plot model for some chi values
```{r}
C_T <- 0.4
chi <- c(10^(-5), 10^(-7), 10^(-9))
k <- waven1[,1]
psi <- matrix(NaN, nrow=length(k), ncol=length(chi))
for(chii in 1:length(chi)){
  psi[,chii] <- C_T * chi[chii] * eps1[1]^(-1/3) * k^(1/3)
}
```

```{r echo=FALSE}
psi <- as.data.frame(psi)
model <- data.frame(k, psi) %>%
  rename(chi1 =V1, chi2 = V2, chi3=V3) %>%
  gather(key = chi, value = psi, -k)

ggplot(model, aes(x = log10(k), y = log10(psi), col=chi)) + geom_line()

```

#### Try MLE fitting

#### Convert measured spectra to wavenumber spectra

```{r}
P_gradT1_c <- matrix(NA, nrow = length(gradT1_c[,1]), ncol=length(gradT1_c[1,]))
for(index in 1:length(gradT1_c[,1])){
  P_gradT1_c[index,] <- gradT1_c[index,] * mean_speed
}
```

##### Choose spectrum to run model on
The sample fit is done with spectrum nr 70, i.e. from around 200 m depth

```{r}
specnr <- 102
segment <- specnr
```

##### define x and y for model, only use spectrum for wavenumbers < 70 cpm
```{r}
eps1 <- eps[1,]
mind <- length(waven1[waven1[,segment]<=10,segment]) # index for subsetting model values
y <- P_gradT1_c[1:mind,segment]   # corrected wavenumber temp-gradient spectrum --> P_gradT1_c
x <- waven1[1:mind,segment] 
```

##### Definition of likelihood function
```{r}
#likelihood function definition
LL <- function(chi) {
  # Find residuals
  R = y - C_T * chi * eps1[segment]^(-1/3) * x^(1/3)
  # Calculate the likelihood for the residuals
  R = suppressWarnings(dnorm(R, log = TRUE))
  # Sum the log likelihoods for all of the data points
  -sum(R)
}

#model fit
fit <- suppressWarnings(mle(LL, start = list( chi=10^(-7))
                            ,nobs = length(y),  method='L-BFGS-B'))
```

##### Results of model fit - chi = `r fit@coef[1]`
```{r echo=FALSE}
summary(fit) 
```


##### calculate model with fitted chi for plotting
```{r}
spec_model <-  C_T * fit@coef[1] * eps1[segment]^(-1/3) * x^(1/3)
```

##### plot data and model
```{r}
pl_data <- data.frame(k_data=waven1[,segment], spec_data=P_gradT1_c[,segment])
pl_model <- data.frame(k_model = x[-1], spec_model = spec_model[-1])
ggplot() + 
  geom_point(data=pl_data, aes(x=k_data, y = spec_data)) + 
  geom_line(data=pl_model, aes(x=k_model, y=spec_model), color='red') +
  scale_y_continuous(trans='log10', limits=c(10^-9,10^-2)) +
  scale_x_log10()
```


